KIN: Method to infer relatedness using low coverage ancient DNA

### 2022-04-04 paper_overleaf/main.tex
Updates from Overleaf

Identifying related individuals is a common task in genetic studies. Relatedness is of direct interest in e.g. DNA forensics, where familial search can aid in solving criminal cases, and to identify unknown deceased persons \cite{murphy_law_2018,ram_genealogy_2018}. Genetic paternity tests have an important application in resolving family relation, e.g. in establishing relationship between a person applying for immigration and the claimed relatives \cite{egeland_beyond_2000}. It is also an essential pre-processing step in population genetics and association studies, where samples are typically assumed to be independent random draws from the population.

@@

\subsection{Approaches to estimate relatedness from high coverage data}

Commonly, pairs of related individuals are identified by looking for parts of the genome that are identical by descent (IBD), ie. inherited from a recent common ancestor. Due to the laws of Mendelian segregation, each parent, for example, will share exactly one set of chromosomes IBD with their offspring. A grandparent will, on average share a quarter of their genome with a grand-child due to recombination. Along the genomes of a pair of diploid individuals, there are three IBD states possible at any given position: either no chromosomes are IBD, only one chromosome is IBD, or both chromosomes are IBD. The genome-wide proportions of these states (usually referred to as $k_0$, $k_1$, $k_2$, so that $k_0+k_1+k_2=1$) can be used to infer the degree and nature of relatedness for a pair of individuals. For example, a pair of siblings are expected to have all three possible IBD states with probabilities (0.25,0.5,0.25) in the absence of inbreeding, as shown in Fig \ref{fig0:schematic}. These IBD probabilities can directly be used to categorize their relatedness as shown in table \ref{tab:Table 1}. One can also use these probabilities to estimate the coefficient of relatedness $r$, which is defined as the proportion of the genome that is IBD. In the absence of inbreeding, this would be calculated as $r= k_1/2 + k_2$.

However, it is not possible to directly observe IBD segments, and so most methods infer them by first identifying segments of the genome that are Identical by State (IBS) and using population allele frequencies to calculate the probability of IBD given IBS \cite{vai_KINship_2020}. One exception is SNPduo, a software that uses IBS directly from a pair of individuals to visualize and analyze their relatedness. \cite{roberson_visualization_2009}. Most methods, however, use IBD as a hidden state in inferring relatedness \cite{boehnke_accurate_1997,lynch_estimation_1999, albrechtsen_natural_2010, purcell_plink_2007}.

@@

These methods utilize population allele frequencies, phase information, recombination maps, or genotype calls to co-estimate IBD and the relatedness coefficient \cite{huff_maximum-likelihood_2011,li_relationship_2014,li_accurate_2014,thornton_estimating_2012}. There are many such methods available. PLINK, a popular tool to infer relatedness coefficient in diploid genotype data, estimates IBD probabilities from observed IBS, using the allele frequencies at each Single Nucleotide Polymorphism (SNP) in the analyzed samples \cite{purcell_plink_2007}. KING is another widely used software that allows relatedness inference with genotype data in homogeneous populations using population allele frequencies in the samples (KING homo), but can also work with structured populations with no good population allele frequencies using heterozygosity estimates from each sample (KING robust) \cite{manichaikul_robust_2010}. .

@@

\subsection{Approaches that address problems with ancient DNA data}
One major issue with applying the above mentioned methods to ancient DNA data is that the sequence coverage is typically low, making it difficult to get genotype calls.
Several methods surmount this problem by using genotype likelihoods \cite{lipatov_maximum_2015,korneliussen_ngsrelate_2015}. In this way, it is possible to account for the uncertainty in genotype calls by summing over all possible genotypes weighted by their genotype likelihoods. However, these approaches typically still require at least 2x coverage, since genotype likelihoods may be unreliable at lower coverages. Ancient DNA analyses often face additional challenges such as contamination from present-day DNA \cite{peyregne_authentict_2020}, absence of a good reference panel, and ascertainment bias in the case of data generated from DNA capture. In cases where the allele frequency or a reference panel for the target population is not available, it is possible to use allele frequency from a modern population from the same geographic location, or to try using population allele frequencies from multiple potentially close populations. However, incorrect assumptions may cause the individuals in the target population to look more similar or different to each other \cite{amorim_understanding_2018}. Several methods have been proposed to estimate relatedness without a reference panel, but many of these require either $>4$x coverage \cite{waples_allele_2019}, or a large sample size to get an estimate of population allele frequencies from the samples \cite{theunert_joint_2017}. In this study, we show that contamination can cause some of the individuals in ancient populations to look more unrelated than they are. This happens because the DNA sequence of an ancient individual typically exhibits more differences to a contaminant sequence than that of another ancient individual from the same population. It is worth mentioning that different DNA libraries may have different levels of contamination, even when they have been sampled from the same specimen. Hence, contamination can reduce the power to detect related pairs, especially when the target population is quite diverged from the contaminating population(s). In addition, the genomes of target individuals may have long runs of homozygosity (ROH) due to a small population size, or recent inbreeding. Long ROH cause related individuals to seem genetically closer to each other, but do not affect the genetic distance between unrelated individuals. This effect may cause an increase in falsely inferred relationships. Moreover, ancient DNA is commonly captured with a SNP array to enrich for sequences from the target species. The way that these SNPs are defined (e.g. selecting SNPs variable in certain individuals) becomes relevant, because methods to infer relatedness based on the fraction of sites in different IBS states are highly sensitive to ascertainment bias, and may give different results when applied to data with different ascertainment schemes \cite{waples_allele_2019}.

Several limitations with ancient DNA mentioned above, such as low coverage, the absence of population allele frequencies or a reference panel, and ascertainment bias are resolved by a software called READ. This software addresses the problem of unavailability of genotype calls by randomly sampling alleles from each individual. A string of these alleles at each position (called pseudo-haploids) are then compared to other individuals to calculate average pairwise genetic distances, which in turn are used to infer relatedness. The limitations are that the software can identify only up to second degree relatives, without differentiating between siblings and parent-child, which may be crucial when making pedigrees.

\subsection{How our method works}
Here, we present KIN (Kinship INference), a Hidden Markov Model (HMM) to estimate relatedness and IBD from low-coverage ancient DNA data. KIN can detect up to \nth{3} degree relatives, and differentiates between siblings and parent-child. KIN is also able to take into account ROH, contamination and ascertainment bias. We validate the performance of KIN using simulations and show that we are able to infer relatedness in real data from two datasets: a group of Neandertals and a group of Bronze age individuals.

@@

\subsection{Model description}\label{method_overview} 
To infer the IBD-state across the genome, we subdivide the genomes of a pair of individuals into $L$ large genomic windows (typically of size 10Mb). The proportion of pairwise differences in those windows will differ based on the IBD-state, and the relatedness case will determine how often we expect to switch between different IBD-states.  The cases we consider are:  Unrelated, $5^{th}$ Degree, $4^{th}$ Degree, $3^{th}$ Degree, Grandparent-Grandchild, Avuncular, Half-siblings, Parent-Child, Siblings, and Identical individuals. The details of the likelihood computation are given in section \ref{ll}. The likelihood computations require an emission (section \ref{B}) and transition probability.  We estimate the transition matrix for each relatedness case from simulations (see section \ref{simulat}), and hence the transition probability for a HMM is fixed. To calculate the emission probabilities, we need to estimate a parameter $\delta$ that describes the variance in the data given the IBD state and the ROH state. The estimation of this parameter is essential to our inference of relatedness, since the variance in the data can change with the number of overlapping sites. So, our algorithm optimizes $\delta$, which is used in the calculation of emission probabilities, and we use these emission probabilities, along with the transition probabilities, to estimate hidden IBD states and to calculate the log likelihood for the HMM. We can then compare the log likelihood from each model to find the best fit along with a measure of uncertainty. We output the IBD states corresponding to the most likely model. 
Hidden IBD state of each window ($Z_w$) is estimated using standard Viterbi algorithm, and $\BZ$ corresponding to the most likely model are returned.

@@

$D_w$ is the number of sites that are different between a pair of individuals, and can be estimated accurately for high coverage data by comparing genotypes. For low-coverage samples, $D_w$ is often calculated by comparing randomly-sampled reads ("pseudo-haploid" sequences) \cite{haak_massive_2015}. However, such an approach may result in loss of data, and hence we calculate $D_w$ by implicitly summing over all possible samplings:
\begin{align}\label{eq:x}
D_w &= \sum_s \nu_i(s) (1-\nu_j(s)) + (1-\nu_i(s)) \nu_j(s)
\end{align}
Here, $\nu_i(s)$ and $\nu_j(s)$ are the proportions of reads carrying the derived allele at SNP index $s$ for individuals i and j respectively.
Throughout, we will use bold-face notation to refer to the vector (or matrix) collecting  all the terms, e.g. $\BD = (D_1, D_2, \dots D_L)$.

\subsection{Log likelihood of the KIN-HMM}\label{ll}
The KIN-HMM uses $\BD$ and $\BN$ to classify each window into three hidden states $Z_w \in$ ($0$, $1$, $2$), reflecting zero, one or two chromosomes IBD. To take ROH into account, we define the variable $H_w \in$ ($0$, $1$, $2$) that designates that zero, one or both individuals are homozygous in window $w$. Since $H_w$ is unobserved, in practice we use the estimates from ROH-HMM $h_{wj} = P(H_w = j)$.

There are three additional model parameters, $\pi, \mathbf{A}$ and $\delta$. The initial probabilities $\pi$ give the probabilities of being in each state $Z$ at the beginning of each chromosome, which we set to uniform for simplicity. The transition matrix $\mathbf{A}$ gives the probability of moving from state $i$ to state $j$, given by $a_{ij}$, and is fixed for each relatedness case. The overdispersion-parameter $\delta$ takes into account that SNPs in each window vary in their allele frequency (see next setion).  For compactness of notation, we group the fixed parameters: $\theta = (\BN, \mathbf{A}, \pi)$.

We compute the complete data likelihood for the HMM:

\begin{align}\label{eq:1}
\log P(\BD,\BZ|\BH, \theta, \delta) &= \log P(\BD|\BZ,\BH, \theta, \delta) + \log P(\BZ |\theta) \nonumber\\
&= \sum_w \log P(D_w|Z_w,H_w, \theta, \delta) + \sum_w \log P(Z_w |Z_{w-1},\theta) + \log P(Z_0|\theta)
\end{align}

@@

Using this set up, we can isolate the emissions $P(D_w | Z_w, H_w ,\theta, \delta)$ from equation \ref{eq:1} and optimize them alone. The number of shared sites $N_w$ is the only parameter in $\theta$ that affects the emissions.

Assuming the sites were equally distributed and independent would result in the binomial likelihood:
$$P(D_w|Z_w, H_w, N_w) \sim \text{Binom}[D_w ; p(Z_w, H_w), N_w] \text{,}$$

where, $p$ is the proportion of differences expected for a particular IBD and ROH state. If the two individuals are unrelated in a particular window (i.e. $Z_w = 0$), then the expected proportion of pairwise differences depends solely on the effective population size, and we denote this proportion with $p_0$. Note that $p_0$ does not depend on the presence of ROH, since this does not change expected proportion of differences between unrelated individuals. If the two individuals share one or even both copies of the genome IBD, we would expect the proportion of differences to be reduced to $p_1 = \frac{3}{4} p_0$, and $p_2 = \frac{1}2 p_0$, respectively, since either one or two of the four possible comparisons will be between identical chromosomes \cite{kuhn_estimating_2018}. Thus, $p(Z_w=i, H_w=0) = p_i$.

The proportion of differences between unrelated individuals $p_0$ is an important parameter. We follow READ \cite{kuhn_estimating_2018} and estimate $p_0$ as the median of differences for all possible pairs of individuals, which should work well if the majority of individuals in the samples are unrelated. We set $p_1$ to $(3/4)p_0$ and $p_2$ to $p_0/2$.

The presence of inbreeding adds an additional complication, as it is possible that all chromosomes are shared between individuals, so that the expected differences should be zero. An example would be inbred regions in two bones from the same individual. However, as we do all our calculations in large windows, the start/end positions of windows may not coincide with that of ROH tracts, and we found that setting expected proportion of differences in this case ($p_3$) at $p_2/2$ works well in practice.

@@

@@

\paragraph{Beta Binomial Model}
The Binomial Model laid out above assumes that sites are independent. However, due to genetic linkage, nearby sites will be correlated and we find that the data often has considerably higher variance than would be expected from a binomial model. We take this into account by adding an overdispersion parameter $\delta$ that allows for variation in the $p$ between windows. Note that, just like $p(Z_w,H_w)$ the overdispersion $\delta(Z_w,H_w)$ depends on the number of chromosomes compared, there are only four parameters $\delta_i$ with $i$ $\in$ (0,1,2,3) analogous to the $p_i$. Empirically, we find that the beta-binomial likelihood provides a good fit to the data (fig. S\ref{figS1:binom})).

Taken together, our emission probabilities are 
\begin{align}\label{eq:3}
P(D_{w}|Z_w,H_w,N_w, \delta) = P(D_{w}|p_i,\delta_i,N_w) &\sim BB[D_w; p_i, \delta_i, N_w] \nonumber\\
&= \binom{N_w}{D_w}\frac{B(D_w+p_i \delta_{i}, N_w-D_w+ \delta_{i}(1-p_{i}))}{ B(p_{i}\delta_{i}, (1-p_{i})\delta_{i})},
\end{align}
where $i$ is fully determined by the combination of $Z_w$ and $H_w$ (see eq 3).

This parameterization of the beta distribution in terms of expected value $p$ and overdispersion $\delta$ is also called the Balding-Nichols-model \cite{balding_method_nodate}, and is distinct from the more common parametrization in terms of $\alpha$ and $\beta$. We use this equation as well if some preprocessing steps result in non-integer $D_w$ and $N_w$, in which case we approximate the binomial coefficient using Gamma functions.

@@

\paragraph{Initialization}
The value of $\delta_i$ is unknown to start with, and we set it to a random value.

@@

\subsection{Contamination correction}\label{contam}
Contamination by the DNA from present-day people is a common feature in ancient DNA \cite{peyregne_present-day_2020}. Even low levels of contamination may make samples look less similar, and thus distort relatedness analyses. This issue is particularly pronounced  when analyzing populations with low diversity, where even $<5\%$ contamination can cause loss of power to detect family relations (Skov et al.,2022 (in press)). To address this issue, we develop a heuristic that adjusts both $D_w$ and $N_w$ to remove contamination from the data.

We conceptualize contamination in the following way. If data from individual $i$ has a known contamination rate of $C_i$, this means that any particular read drawn from this individual will actually come from that individual with probability $1-C_i$. With probability $C_i$ the read comes from a contaminant. Thus, when comparing two reads, one drawn from individual $i$ and the other drawn from individual $j$, there are three possible cases: First, with probability $(1-C_i)(1-C_j)$, both reads are endogenous and we can use them to infer relatedness. With probability $C_i(1-C_j) + C_j(1-C_i)$, one read is endogenous and the other is contaminant. Finally, with probability $C_iC_j$, both reads are contaminant. If we make the simplifying assumption that both $C_i$ and $C_j$ are small, the terms of the order of $C^2$ can be ignored. Then, we compare contaminant read to endogenous read with the probability $C_{ij} = C_i + C_j$.

Given that we know the level of contamination (estimated from other methods),our goal then, is to estimate the expected number of differences from comparison of endogenous reads $\mathbb{E}[D_w^{'}]$, and the total number of sites with overlapping endogenous reads $\mathbb{E}[N_w^{'}]$. To estimate $\mathbb{E}[D_w^{'}]$, we first need to calculate $P(c=0|d=1)$. Here, $c$ is a binary variable with value $0$ or $1$ denoting that the comparison at a site is between endogenous reads, or between endogenous and contaminant reads respectively. Similarly $d$ is a binary variable taking values 0 and 1 for seeing no difference, or a difference, when comparing the reads at a given site. We can further expand this probability:

@@

>>>>

Identifying related individuals is a common task in genetic studies. Relatedness is of direct interest in e.g. DNA forensics, where familial search can aid in solving criminal cases, and to identify unknown deceased persons \cite{murphy_law_2018,ram_genealogy_2018}. Genetic paternity tests have an important application in resolving family relation, e.g. in establishing relationship between a person applying for immigration and the claimed relatives \cite{egeland_beyond_2000}. It is also an essential  step in population genetics and association studies, where samples are typically assumed to be independent random draws from the population.

@@

\subsection{Approaches to estimate relatedness from high-coverage data}

Commonly, pairs of related individuals are identified by looking for parts of the genome that are identical by descent (IBD), ie. inherited from a recent common ancestor. Due to the laws of Mendelian segregation, each parent, for example, will share exactly one set of chromosomes IBD with their offspring. A grandparent will, on average share a quarter of their genome with a grand-child due to recombination. Along the genomes of a pair of diploid individuals, there are three IBD states possible at any given position: either no chromosomes are IBD, only one chromosome is IBD, or both chromosomes are IBD. The genome-wide proportions of these states (usually referred to as $k_0$, $k_1$, $k_2$, so that $k_0+k_1+k_2=1$) can be used to infer the degree and nature of relatedness for a pair of individuals. For example, a pair of siblings are expected to have all three possible IBD states with proportions of 0.25,0.5,0.25, respectively (Fig \ref{fig0:schematic}). These IBD probabilities can directly be used to categorize their relatedness as shown in table \ref{tab:Table 1}. One can also use these probabilities to estimate the coefficient of relatedness $r$, which is defined as the proportion of the genome that is IBD. In the absence of inbreeding, this would be calculated as $r= k_1/2 + k_2$.

However, it is not possible to directly observe IBD segments, and so a common approach is to first identify segments of the genome that are Identical by State (IBS) and using population allele frequencies obtained from a reference panel to calculate the probability of IBD given IBS \cite{vai_KINship_2020, boehnke_accurate_1997,lynch_estimation_1999, albrechtsen_natural_2010, purcell_plink_2007}. One exception is SNPduo, a software that uses IBS directly from a pair of individuals to visualize and analyze their relatedness \cite{roberson_visualization_2009}.

@@

These methods utilize population allele frequencies, phase information, recombination maps, or genotype calls to co-estimate IBD and the relatedness coefficient \cite{huff_maximum-likelihood_2011,li_relationship_2014,li_accurate_2014,thornton_estimating_2012}.  PLINK, a popular tool to infer relatedness coefficient in diploid genotype data, estimates IBD probabilities from observed IBS, using the allele frequencies at each Single Nucleotide Polymorphism (SNP) in the analyzed samples \cite{purcell_plink_2007}. KING is another widely used software that allows relatedness inference with genotype data in homogeneous populations using population allele frequencies in the samples (KING homo), but can also work with structured populations with no good population allele frequencies using heterozygosity estimates from each sample (KING robust) \cite{manichaikul_robust_2010}.

@@

\subsection{Approaches that address problems with ancient DNA data}
One major issue with applying the above mentioned methods to ancient DNA data is that the sequence coverage is typically low, making it difficult to get genotype calls.
Several methods surmount this problem by using genotype likelihoods \cite{lipatov_maximum_2015,korneliussen_ngsrelate_2015}. In this way, it is possible to account for the uncertainty in genotype calls by summing over all possible genotypes weighted by their genotype likelihoods. However, these approaches typically still require at least 2x coverage, since genotype likelihoods may be imprecise at lower coverages . Ancient DNA analyses often face additional challenges such as contamination from present-day DNA \cite{peyregne_authentict_2020}, absence of suitable reference panels, and ascertainment bias in the case of data generated from DNA capture.

In cases where the allele frequency or a reference panel for the target population is not available, it is possible to use allele frequency from a modern population from the same geographic location, or to try using population allele frequencies from multiple potentially close populations. However, incorrect assumptions may cause the individuals in the target population to look more similar or different to each other \cite{amorim_understanding_2018}. Several methods have been proposed to estimate relatedness without a reference panel, but many of these require either $>4$x coverage \cite{waples_allele_2019}, or a large sample size to get an estimate of population allele frequencies from the samples \cite{theunert_joint_2017}. In this study, we show that contamination can cause some of the individuals in ancient populations to look more unrelated than they are. This happens because the DNA sequence of an ancient individual typically exhibits more differences to a contaminant sequence than that of another ancient individual from the same population. It is worth mentioning that different DNA libraries may have different levels of contamination, even when they have been sampled from the same specimen. In addition, some analyzed  genomes  may have long runs of homozygosity (ROH), for example due to a small population size, or due to recent inbreeding. Long ROH cause related individuals to seem genetically more similar to each other, but do not affect the genetic distance between unrelated individuals.

Moreover, ancient DNA is commonly captured with a SNP array to enrich for sequences from the target species. Particularly methods based  on the fraction of sites in different IBS states are sensiteive to the ascertainment bias caused by this non-random selection of targted sites \cite{waples_allele_2019}.

Several limitations with ancient DNA mentioned above, such as low coverage, the absence of population allele frequencies or a reference panel, and ascertainment bias are resolved by a software called READ (REF). This software addresses the problem of unavailability of genotype calls by randomly sampling alleles from each individual. A string of these alleles at each position (called pseudo-haploids) are then compared to other individuals to calculate average pairwise genetic distances, which in turn are used to infer relatedness. However, READ only infers the degree of relatedness, and only up to second degree.

\subsection{How our method works}
Here, we present KIN (Kinship INference), a Hidden Markov Model (HMM)-based approach to estimate relatedness and IBD from low-coverage ancient DNA data. KIN can detect up to \nth{3}-degree relatives, and differentiates between siblings and parent-child. KIN is also able to take into account ROH, contamination and ascertainment. We validate the performance of KIN using simulations and show that we are able to infer relatedness in real data from two datasets: a group of Neandertals and a group of Bronze age individuals.

@@

\subsection{Model description}\label{method_overview}

The goal of KIN-HMM is to infer how two individuals are  relateded via the patterns of shared IBD-states along a pair of genomes. For this purpose,  we subdivide the genomes of a pair of individuals into $L$ large genomic windows (typically of size 10Mb), and infer the pattern of IBD-sharing for each relatedness case we consider (Unrelated, $5^{th}$ Degree, $4^{th}$ Degree, $3^{th}$ Degree, Grandparent-Grandchild, Avuncular, Half-siblings, Parent-Child, Siblings, and Identical). We then compare the likelihood between all models, and classify each pair to the model with the highest likelihood. We also return the most likely location of IBD tracts (using standard Viterbi algorithm), and the state posterior probablities $\BZ$.

The details of the likelihood computation are given in section \ref{ll}. As any HMM, KIN-HMM  require a set of n emission (section \ref{B}) and transition probability matrices. We assume the transition matrix for each relatedness case is fixed. Our emissions include a vector of parameters ($\delta$) that describes the variance in the data for each IBD state that we infer from the data (section ADD REF).

@@

$D_w$ is the number of sites that are different between a pair of individuals. For high-coverage data, this can be estimated accurately by comparing genotypes, but for low-coverage samples, $D_w$ needs to be estimated. The simplest approach is is to  randomly-sampled a read from each position ("pseudo-haploid" sequences) \cite{haak_massive_2015}. However, such an approach may result in loss of data, and hence we calculate $D_w$ by implicitly summing over all possible samplings:
\begin{align}\label{eq:x}
D_w &= \sum_{s=1}^{N_w} \nu_i(s) (1-\nu_j(s)) + (1-\nu_i(s)) \nu_j(s)
\end{align}
Here, $\nu_i(s)$ and $\nu_j(s)$ are the proportions of reads carrying the derived allele at SNP index $s$ for individuals i and j respectively.
Throughout, we will use bold-face notation to refer to the vector (or matrix) collecting  all the terms, e.g. $\BD = (D_1, D_2, \dots D_L)$.

\subsection{Log likelihood of the KIN-HMM}\label{ll}
The KIN-HMM uses $\BD$ and $\BN$ to classify each window into three hidden states $Z_w \in$ ($0$, $1$, $2$), reflecting zero, one or two shared chromosomes IBD, respectively. To take ROH into account, we define the variable $H_w \in$ ($0$, $1$, $2$) that designates that zero, one or both individuals are homozygous in window $w$. Since $H_w$ is unobserved, in practice we use the estimates from ROH-HMM $h_{wj} = P(H_w = j)$.

There are three additional model parameters, $\pi, \mathbf{A}$ and $\delta$. The initial probabilities $\pi$ give the probabilities of being in each state $Z$ at the beginning of each chromosome, which we set to uniform for simplicity. The transition matrix $\mathbf{A}$ gives the probability of moving from state $i$ to state $j$, given by $a_{ij}$, and is fixed and estimated from simulations for each relatedness case. The overdispersion-parameter $\delta$ takes into account that SNPs in each window vary in their allele frequency (see next setion).  For compactness of notation, we group the fixed parameters: $\theta = (\BN, \mathbf{A}, \pi)$.

Thus, the complete data likelihood for the HMM is

\begin{align}\label{eq:1}
\log P(\BD,\BZ|\BH, \theta, \delta) &= \log P(\BD|\BZ,\BH, \theta, \delta) + \log P(\BZ |\theta) \nonumber\\
&= \sum_w \log P(D_w|Z_w,H_w, N_w, \delta) + \sum_w \log P(Z_w |Z_{w-1},A) + \log P(Z_0|\pi)\text{.}
\end{align}

@@

Using this setup, we can isolate the emissions $P(D_w | Z_w, H_w ,\theta, \delta)$ from equation \ref{eq:1} and optimize them for $\delta$. The simplest model is to assume that sites in each window are equally distributed and independent, in this case we could use the  binomial likelihood:
$$P(D_w|Z_w, H_w, N_w) \sim \text{Binom}[D_w ; p(Z_w, H_w), N_w] \text{,}$$

where, $p$ is the proportion of differences expected for a particular IBD and ROH state. If the two individuals are unrelated in a particular window (i.e. $Z_w = 0$), then the expected proportion of pairwise differences depends solely on the population history, and we denote this proportion with $p_0$. If the two individuals share one or even both copies of the genome IBD, we would expect the proportion of differences to be reduced to $p_1 = \frac{3}{4} p_0$, and $p_2 = \frac{1}2 p_0$, respectively, since either one or two of the four possible comparisons will be between identical chromosomes \cite{kuhn_estimating_2018}. Thus, $p(Z_w=i, H_w=0) = p_i$.

The proportion of differences between unrelated individuals $p_0$ is an important parameter. We follow READ \cite{kuhn_estimating_2018} and estimate $p_0$ as the median of differences for all possible pairs of individuals, which should work well if the majority of individuals in the samples are unrelated. We set $p_1$ to $(3/4)p_0$ and $p_2$ to $p_0/2$, because one, respectively two of the four compared chromosomes are identical in these cases.

The presence of inbreeding adds an additional complication, as the number of shared chromosome may  change. For example, when considering two bones from the same individual, we would expect the entire genome to have a pairwise difference of $p_2$, because two out of the four compared chromosomes are between identical copies. However, in inbred regions all four chromosomes will be identical, and so the expected pairwise differences are zero. Note however, that $p_0$ does not depend on the presence of ROH, since all comparisons are between unrelated chromosomes even if both individuals are homozygous at a particular locus.

@@

As explained above, $p_3 = 0$. However, as we do all our calculations in large windows, the start/end positions of windows may not coincide with that of ROH tracts, and we found that we obtain better results by setting $p_3 = \frac{p_2}{2}$).

@@

\paragraph{Beta Binomial Model}
The Binomial Model laid out above assumes that sites are independent. However, due to genetic linkage, nearby sites will be correlated. As a result, we empirically find that the data often has considerably higher variance than would be expected from a binomial model (fig. S\ref{figS1:binom}). We take this into account by adding an overdispersion parameter $\delta$.Just like $p(Z_w,H_w)$, $\delta(Z_w,H_w)$ depends on the number of chromosomes compared and so each of the  four $p_i$ has a corresponding $\delta_i$ parameter

Taken together, our emission probabilities are 
\begin{align}\label{eq:3}
P(D_{w}|Z_w,H_w,N_w, \delta) &\sim BB[D_w; p_i, \delta_i, N_w] \nonumber\\
&= \binom{N_w}{D_w}\frac{B(D_w+p_i \delta_{i}, N_w-D_w+ \delta_{i}(1-p_{i}))}{ B(p_{i}\delta_{i}, (1-p_{i})\delta_{i})},
\end{align}
where $i$ is fully determined by the combination of $Z_w$ and $H_w$ (see eq 3).

This parameterization of the beta distribution in terms of expected value $p$ and overdispersion $\delta$ is also called the Balding-Nichols-model \cite{balding_method_nodate}, and is distinct from the more common parametrization in terms of $\alpha$ and $\beta$. We use this equation even if preprocessing steps result in non-integer $D_w$ and $N_w$, in which case we approximate the binomial coefficient using Gamma functions.

@@

\paragraph{Initialization}
The value of $\delta_i$ is unknown to start with, and we set it to a random value between x and y.

@@

\subsection{Contamination correction}\label{contam}
Contamination by the DNA from present-day people is a common feature in human ancient DNA \cite{peyregne_present-day_2020}. Even levels of contamination may make samples look less similar, and thus distort relatedness analyses. This issue is particularly pronounced  when analyzing populations with low diversity, where even $<5\%$ contamination can cause loss of power to detect family relations . To address this issue, we develop a heuristic that adjusts both $D_w$ and $N_w$ to remove contamination from the data.

We conceptualize contamination in the following way: If data from individual $i$ has a known contamination rate of $C_i$, this means that any particular read  will actually come from the desired individual with probability $1-C_i$. With probability $C_i$ the read comes from a contaminant. Thus, when comparing two reads, one drawn from individual $i$ and the other drawn from individual $j$, there are three possible cases: First, with probability $(1-C_i)(1-C_j)$, both reads are endogenous and we can use them to infer relatedness. With probability $C_i(1-C_j) + C_j(1-C_i)$, one read is endogenous and the other is contaminant. Finally, with probability $C_iC_j$, both reads are contaminant. If we make the simplifying assumption that both $C_i$ and $C_j$ are small, the $C_iC_j$ terms  can be ignored. Then, we compare contaminant read to endogenous read with the probability $C_{ij} = C_i + C_j$.

We assume that the level of contamination is known (cite Stephane and Kay's review). We  estimate the expected number of differences from comparison of endogenous reads $\mathbb{E}[D_w^{'}]$, and the total number of sites with overlapping endogenous reads $\mathbb{E}[N_w^{'}]$. To estimate $\mathbb{E}[D_w^{'}]$, we first need to calculate $P(c=0|d=1)$. Here, $c$ is a binary variable with value $0$ or $1$ denoting that the comparison at a site is between endogenous reads, or between endogenous and contaminant reads respectively. Similarly $d$ is a binary variable taking values 0 and 1 for seeing no difference, or a difference, when comparing the reads at a given site. We can further expand this probability:

@@

%adding back the previous versions:
If we look at one site in the window, we could find either no difference in reads from individuals $i$ and $j$ ($d_s$=0), or a non-zero difference ($0<d_s<=1$). Here $d_s$ is the probability of seeing a difference at site $s$, and $\sum_{s \in w} d_s =D_w$. We ask, what is the

If we make the simplifying assumption that both $C_i$ and $C_j$ are small, terms of the order of $C^2$ can be ignored and there are just two cases: with probability $C_{i,j} =C_i + C_j$, one of the reads stems from a contaminant, and we effectively compare a contaminant fragment with an endogenous one. The probability that they differ is the average divergence between target and contaminating population, which we denote with $\phi$. The other case (w. p. $(1-C_{i,j}$) is where we compare two endogenous reads, which will differ with probability

\begin{equation}
    p_{e} = \frac{p_{obs}-\phi C_{i,j}}{1-C_{i,j}}
\end{equation}

Here, $$p_{obs} = \frac{\sum_w D_w}{\sum_w N_w}$$

We estimate $\hat{D_c} = C_{i,j} N_{tot} \phi$ and $\hat{D_e} = (1-C_{i,j}) N_{tot} p_{obs}\phi$.
If there are $N_{obs}$ overlapping sites in a window, $N_{c} = C_{i,j}N_{obs}$ of them will include a contaminant allele, and $N_e = (1-C_{i,j})N_{obs}$ will be between endogenous alleles.

For KIN, we can use $N_e$ as an input (for each window), but we also need the proportion of differences, $D_e$. For this, we use the assumption that the proportion of sites where the contaminant differs from our samples is

$$\phi = \frac{D_c}{N_c} = \frac{D_c}{C_{i,j} N_{obs}}, $$
and thus $D_c = C_{i,j} \phi N_{obs}$.

We can think of $p_c$ and $\rho$ as empirical likelihoods that a particular site shows a difference between two reads, given that the comparisons contain only endogenous or one contaminant read, respectively. 
\begin{align}
    P(E|D) &= \frac{P(D|E)P(E)}{P(D)} \nonumber\\
    \mathbb{E}[D_e] &= D_{obs} P(E|D) = D_{obs}\frac{p_e (1-c)}{p_e(1-c) + c\rho}\\
    \mathbb{E}[S_e] &= D_{obs} P(E|\text{not} D) = S_{obs}\frac{(1-p_e)(1-c)}{(1-p_e)(1-c) + c(1-\rho)}
\end{align}

And for calculating $N_{e}$, we added $D_{e}$ to $S_{e}$

The goal is that given $C_{i,j}, p_e, \BN and \BD$ to calculate adjusted $\hat{\BN}$ and $\hat{\BD}$ that reflect the endogenous proportion only.

, and the observed proportion of differences between individuals $i$ and $j$ as $p_{oij}$. We calculate average endogenous pairwise difference $p_{eij}$ for individuals i and j as follows \cite{noauthor_ancient_nodate}:
\begin{align}
    p_{oij} = C_{ij}  p_c + (1-C_{ij})  p_{eij}\\
    or, p_{eij} = \frac{p_{oij} - C_{ij}  p_c}{1-C_{ij}} 
\end{align}

Given $p_{eij}$, we would ideally like to calculate the complete data log likelihood for the relatedness model while summing over all possible differences we could have in a window weighted by the probability of seeing that difference.

\paragraph{Short version}
With contamination, we do not observe the number of differences $D_{true}$ and total number of comparisons $N_{true}$ directly, but instead need to approximate them. The window-subscript is omitted in the following for clarity:
\begin{equation}
    P(D_{\text{obs}} | N_\text{obs}, Z, \theta) \approx 
    P( \hat{D} | \hat{N}, Z, \theta)
\end{equation}
We do that by calculating the probability that a single read is endogenous given it has a difference.
\begin{align}
    \hat{D} = \mathbb{E}[D | D_{\text{obs}}, c] &= 
    D_{\text{obs}} P(E_i | D_i=1) \\
    &= D_\text{obs} \frac{P(D_i=1 | E_i)P(E_i)}{P(D_i=1)}\\
    &= D_\text{obs} \frac{(1-c)p_e}{(1-c)p_e + c p_c}
\end{align}

and likewise for reads that do not have a difference
\begin{align}
    \hat{S} = \mathbb{E}[S | S_{\text{obs}}, c] &= 
     S_\text{obs} \frac{(1-c)(1-p_e)}{(1-c)(1-p_e) + c (1-p_c)}
\end{align}

and $\hat{N} = \hat{S} + \hat{D}$

\begin{equation}
    P(D_{\text{obs}} | Z, N_\text{obs}, \theta) = \sum_n\sum_d P(D_{\text{obs}}, N_\text{obs} | D_w=d, N_w, C_i, C_j)\underbrace{P(D_w=d | N_w, Z, \theta)}_\text{old likelihood}
\end{equation}

i.e. we sum over all possible $D_w$, $N_w$. Instead, we use

\begin{align}
    P(D_{cor},Z|D,H,\theta,c,p_c) &= P(D_{cor}|Z,D,H,\theta,c,p_c) P(Z|D,H,\theta,c,p_c)\nonumber\\
    &= [\prod_{w} [\sum_\kappa P(D_{cor,w}|Z_w, H_w, \theta,c,p_c) P(D_{cor,w}=\kappa)] \prod_{w} P(Z_w|Z_{w-1}, \theta)] P(Z_0| \theta)
\end{align}

We realize that this calculation would be time consuming, and would further complicate the HMM. To avoid this, instead of using weighted sum of all possible $\kappa$, we use $E(D_{cor,w}|Z_w, H_w, \theta,c,p_c)$. This simplifies the equation 21 to :
